Liveness-Based Garbage Collection for Lazy Languages

### 2015-04-28 paper/paper.tex
Updated paper as per discussion

We next  describe the function $\mathcal{L}$  that propagates liveness
across  expressions.   This brings  us  to  the  second way  in  which
liveness analysis for lazy languages differ from eager languages---the
point  where  an  expression   gets  evaluated  cannot  be  statically
determined.  To  deal with this, we introduce an analysis which is
flow-insensitive but path-sensitive\comment{Check the terminology} into
our  analysis.

\begin{figure*}[t!]
\begin{picture}(100,130)(0, -50)
  \begin{tabular}{cc}
    \begin{boxedminipage}{.51\textwidth}
      {\sf
        \small
        \renewcommand{\arraystretch}{1}{
          \begin{uprogram}
            \UNL{1} (\DEFINE\ (\length~\xl)
            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\ $\psi_1\!\!:\,$ \px
            \UNL{4} \hspace*{1.15cm} $\psi_2\!\!:\, (\RETURN ~0)$
            \UNL{4} \hspace*{1.15cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
            \UNL{5} \hspace*{1.15cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
            \UNL{6} \hspace*{1.15cm} $\pi_7\!\!:\,
            $(\LET~\pz\ $\leftarrow$ (1~+~\py))))))) \IN
            \UNL{7} \hspace*{1.15cm} $\psi_3\!\!:\, (\RETURN ~\pz)$)))))
        \end{uprogram}}
        \renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} $\pi_\mainpgm\!\!:\, \pi_8\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
	  \UNL{2}   \hspace*{.78cm}   $\pi_9\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
          \UNL{3}   \hspace*{.78cm}    $\pi_{10}\!\!:\,
          $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN
          \UNL{4}  \hspace*{.78cm}  $\psi_{4}\!\!:\,$ (\RETURN\ \pw)))))))
\end{uprogram}}
        } 
    \end{boxedminipage}
    & 
    \begin{boxedminipage}{.49\textwidth}
      {\sf
        \small
        \begin{tabular}{lc}
        $\cal{M_{\mbox{\tiny \length}}} = $ 
            \begin{tabular}{lc}            
              $[$
              $ep_2 \mapsto \lbrace \py.\epsilon, \pu.\Lf{\length}{\mbox{1}}{\epsilon}, \
              \xl.\mbox{1}\Lf{\length}{\acdr}{\epsilon}, \xl.\epsilon  \rbrace ,$ \\ 
              $ep_1 \mapsto \lbrace \rbrace , $\\
              $ep_3 \mapsto \lbrace \px.\epsilon, \xl.\epsilon \rbrace $  \\
              $]$
            \end{tabular} \\
            $\cal{P_{\mbox{\tiny \length}}} = [$
              \begin{tabular}{lc}
                $\pi_{\mbox{1}} \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \
                $ \pi_{\mbox{2}} \mapsto \lbrace ep_1, ep_2, ep_3 \rbrace, $ \\
                $ \pi_{\mbox{3}} \mapsto \lbrace ep_3 \rbrace, $ \
                $ \pi_{\mbox{4}} \mapsto \lbrace ep_1 \rbrace, $ \
                $ \pi_{\mbox{5}} \mapsto \lbrace ep_2 \rbrace, $ \\
                $ \pi_{\mbox{6}} \mapsto \lbrace ep_2 \rbrace, $ \
                $ \pi_{\mbox{7}} \mapsto \lbrace ep_2 \rbrace $ \\
                $]$
              \end{tabular}\\
              $\cal{M_{\mbox{\tiny \main}}} = $ 
              \begin{tabular}{lc}            
                $[$
                  $ep_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}, \pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace $ \
                  $]$
              \end{tabular} \\
              $\cal{P_{\mbox{\tiny \main}}} = [$
                \begin{tabular}{lc}
                  $\pi_{\mbox{8}} \mapsto \lbrace ep_4 \rbrace, $ \
                  $ \pi_{\mbox{9}} \mapsto \lbrace ep_4 \rbrace, $ \\
                  $ \pi_{\mbox{10}} \mapsto \lbrace ep_4 \rbrace, $ \
                  $ \pi_{\mbox{11}} \mapsto \lbrace ep_4 \rbrace, $ \
                  $ \ldots  $ \
                  $]$  
                \end{tabular}
        \end{tabular}
      }
    \end{boxedminipage}
    \\
    (a) Example program.&
    \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
        (b) {\em e-path} to liveness environment map \\
        (c) Prog pt to {\em e-paths } map
    \end{tabular}}
  \end{tabular}
\end{picture}
\kern -3ex
\vspace{0.5in}
\caption{Example Program and its liveness maps.}\label{fig:mot-example2}    
\end{figure*}

@@

Since   \pz\  is  $(1 + ~\py)$,  the   liveness  of   $\sigma$  for
\pz\ translates  into a demand  of $\sigma$ on $(1+~\py)$,  which in
turn  generates  a  liveness  of $\{\py.\epsilon\}$  at
$\pi_7$.  The liveness of \py\ should be killed at $\pi_6$ since it is
beyond the scope of the definition of \py.  However, this does not cause any imprecision
in the way of the GC marking  more cells as live, since \py\ is yet to
come into existence at $\pi_6$  and its liveness will not be consulted
by the garbage collector at this point.
In an eager language, \py\ would not have been live at $\psi_3$, the
point of  its last  use. In a  lazy language, however,  the definition
$\pz  \leftarrow (1+~\py)$,  does  not result  in  an evaluation  of
$(1+~\py)$; instead  a closure is created.   The evaluation actually
takes place while evaluating  $(\RETURN~\pz)$. Thus it is necessary to
record the liveness  of $\py$ at the program  point $\psi_3$, indeed at
every program  point from $\psi_3$ up  to the beginning  of the program
(since  we do  not kill  liveness).  We  thus regard  the body  of the
function as  a tree and consider paths  from the root of  this tree to
program  points  which  trigger  a evaluation\footnote{Program points
  which trigger evaluation are labelled with $\psi$}---\SIF\  conditions  and
\RETURN\  expressions.  We  call such  paths {\em
  e-paths}. \comment{Update the e-path labels}  For the
program shown, there are three such paths $EP_1 = \pi_1, \pi_2, \pi_3,
\pi_4,  \pi_5$, $EP_2  =  \pi_1, \pi_2,  \pi_3,  \pi_4, \pi_6,  \pi_7,
\pi_8$, and $EP_3 = \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.

The analysis  consists of  two maps: The  map $\cal{M}$ maps  a e-path to a common liveness environment, and a second map $\cal{P}$ maps
a program  point to the set of  e-paths on which  the point lies.
If a  program  point happens to fall  on more than one  such path, then
the  liveness environment of  the program  point is  the variable-wise
union of  the liveness environments  of the individual paths.   In the
example program, ${\cal M}(\psi_1)= \ldots$, ${\cal M}(\psi_2)=\ldots$ and
${\cal M}(\psi_3)=  \ldots$.  Since  ${\cal P}(pi_4)$ is  $\lbrace EP_1,
EP_2, EP_3 \rbrace$, the liveness  of $pi_4$ is \ldots.  In summary,
our propagation of liveness is flow-insensitive but path-sensitive.
In the sequel
we  shall  describe  how  to  obtain $\cal{M}$.   The  formulation  of
$\cal{P}$ is easy and we do not elaborate it any further.

@@

\section{Generating Liveness Equations}
The  program  in  Figure~\ref{fig:mot-example2}  serves as  a  running
example. Given  an arbitrary function $\mathit{f}$,  we now illustrate
how  to generate  the equations  describing the  demand transformation
\Lfonly$_\mathit{f}$.   Starting with a  symbolic demand  $\sigma$, we
determine   \Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.    Of   particular
interest  is the liveness  at the  program point  at the  beginning of
$e_{\mathit{f}}$ (assumed  to be $\pi_1$).   We generate
equations     of     the     form    $\Lf{\mathit{f}}{i}{\sigma}     =
\Lanv{\pi_1}{x_i}$, where  $x_i$ is the  $i^{\mbox{\footnotesize th}}$
parameter of $\mathit{f}$. Doing this for \length, we get:
\begin{align*}
 \Lanv{\pi_1}{\xl} &= \cbar\sigma \cup \acdr\Lf{\length}{1}{\sigma}
\end{align*}
Which gives the following equation for \Lfone{\length}: 
  \begin{eqnarray*}
    \Lf{\length}{1}{\sigma}
    &=& \cbar\sigma \cup \acdr\Lf{\length}{1}{\sigma}
   \end{eqnarray*}

@@

As mentioned earlier\comment{Check if already mentioned}, each of  the liveness rules modifies a demand by
prefixing   it with   symbols   in   the   above   set,   and   therefore
$\Lf{f}{i}{\sigma}$ has the closed form:
\begin{eqnarray}
\label{eq:LF:DI}
  \Lf{f}{i}{\sigma} &=& \Df{f}{i}.\sigma
\end{eqnarray}
where \Df{f}{i} are sets of  strings over the alphabet $\lbrace \acar,
\acdr,\bcar, \bcdr, \cbar \rbrace$.   Substituting the guessed form in
the  equation  describing  \Lfonly$_{\mathit  f}$, and  factoring  out
$\sigma$,  we  get an  equation  for  \Df{f}{i} that is
independent of $\sigma$.  Any solution for \Df{f}{i} yields a solution
for \Lfonly$_{\mathit f}$.  Applied to \Lfonly$_{\mathit \length}$, we
get:
  \begin{align*}
    \Df{\length}{1} &= \lbrace\cbar\rbrace 
    \cup \acdr\Df{\length}{1}
  \end{align*}

@@

The expression $e_{\mainpgm}$  is a special case; we  assume it may be
called externally with demand $\sigma_\mainpgm = \{\acar,\acdr\}^\ast$
(denoted $\sigma_{\!all}$).  This is because any part of its value may
be used by the environment---for printing the result, for instance.
{\color {red} What about print?}
For the running example, $\length$ has calls from $\main$ at $\pi_{11}$
and a recursive call at $\pi_6$.
%\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
So $\sigma_{\length} =
     \deltacall{\length}{\pi_{11}}{\main}  \cup \deltacall{length}{\pi_6}{\length}$.
Filling in  the values gives:

\begin{align*}
\sigma_{\length}    &=
 \sigma_{\!all}  ~\cup~ \acdr\Lf{length}{1}{\sigma_{\mathit{\length}}}
\end{align*}

For some of the program points in \length\  and \main,  this gives
{\color {red} Prasanna, please fill this correctly}. You could use the
set notation if the map notation spills over: 
\begin{align*}
\Lan{\pi_1} &= \Lan{\pi_2} = \ldots \\
\Lan{\pi_3} &= \ldots \\
\Lan{\pi_4} &= \ldots \\
\Lan{\pi_5} &= \Lan{\pi_6} = \Lan{\pi_7} \ldots \\
\Lan{\pi_8} &= \Lan{\pi_9} = \Lan{\pi_{10}} \ldots
\end{align*}

In summary, the equations generated during liveness analysis are:
\begin{enumerate}
\item   For    each   function   $\mathit{f}$,    equations   defining
\Df{f}{i} for use by \Lfonly$_{\mathit f}$.
\item  For  each   function  $\mathit{f}$, an  equation  defining
the summary demand  $\sigma_{\mathit f}$ on $e_f$.
% \marginpar{Consistency: summary vs cumulative?}
\item For each function $\mathit{f}$ (including $\mainpgm$ for
  $e_\mainpgm$) \comment{Have we defined GC points for lazy languages?}
an equation defining
  liveness at each GC point of $e_{\mathit f}$.

\subsection{Solving liveness equations---the grammar interpretation}
The  equations above can now be re-interpreted as a
context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
\bcar, \bcdr, \cbar\rbrace$.  Let  \var{$X$} denote the non-terminal for
a variable  $X$ occurring on  the LHS of the  equations generated
from the analysis.  We can  think of the resulting productions as
being associated with several  grammars, one for each non-terminal
\var{\Lanv{i}{x}} regarded as a start symbol.  As an example, the
grammar    for   \var{\Lanv{7}{\xl}}   comprises    the   following
productions:
 \begin{eqnarray*}
\var{\Lanv{7}{\xl}}  &\rightarrow&   \var{\acdr \Lf
  {\length}{1}{\epsilon}} \mid \epsilon \\
\end{eqnarray*}
\comment{Pick up a richer equation or more than one liveness equations}
Other equations can be converted  similarly.     The  language  generated  by  \var{\Lanv{i}{x}},
denoted $\mathscr{L}(\var{\Lanv{i}{x}})
$,  is the desired  solution of
\Lanv{i}{x}.   
However, recall from  our earlier  discussion that
the  decision problem that  we are  interested in  during garbage
collection is:
\begin{quote}
Let $x.\alpha$ be  a forward  access path  consisting  only  of   edges
\acar\  and \acdr\  (but  not \bcar\  or  \bcdr).   Let              $\mathscr{L}(\var{\Lanv{i}{x}})
\stackrel{*}{\hookrightarrow} \sigma$, where  $\sigma$  consists 
of forward paths only. Then does $\alpha \in \sigma$?
\end{quote}
We could convert  the  rules  defining  $\hookrightarrow$  into
productions and  add them to the grammar. However, this results
in  an {\em  unrestricted grammar}~\cite{hopcraft90toc},  and the
membership  problem   for  such  grammars   is
undecidable. \comment{Rewrite it to say that this is a CFG with 3
  unrestricted rules and the membership for even this grammar is undecidable}
\comment{State the undecidability of the decision problem above.}

>>>>

We next  describe the function $\mathcal{L}$  that propagates liveness
across  expressions.   This brings  us  to  the  second way  in  which
liveness analysis for lazy languages differ from eager languages---the
point  where  an  expression   gets  evaluated  cannot  be  statically
determined.  To  deal with this, we introduce an analysis which is
flow-insensitive but path-sensitive\comment{Checked the terminology, path-sensitive means certain prooperties holding along a path only, http://en.wikipedia.org/wiki/Data-flow\_analysis} into
our  analysis.

\begin{figure*}[t!]
\begin{picture}(100,130)(0, -50)
  \begin{tabular}{cc}
    \begin{boxedminipage}{.51\textwidth}
      {\sf
        \small
        \renewcommand{\arraystretch}{1}{
          \begin{uprogram}
            \UNL{1} (\DEFINE\ (\length~\xl)
            \UNL{2}  $\pi_1\!\!:\, $(\LET\ \px\ $\leftarrow $\ (\NULLQ~\xl) \IN
            \UNL{3} \hspace*{.05cm} $\pi_2\!\!:\,$(\SIF\ $\psi_1\!\!:\,$ \px
            \UNL{4} \hspace*{1.15cm} $\psi_2\!\!:\, (\RETURN ~0)$
            \UNL{4} \hspace*{1.15cm}    $\pi_5\!\!:\, $(\LET~\pu\  $\leftarrow$  (\CDR~\xl)  \IN
            \UNL{5} \hspace*{1.15cm}   $\pi_6\!\!:\, $(\LET~\py\  $\leftarrow$  (\length~\pu)  \IN
            \UNL{6} \hspace*{1.15cm} $\pi_7\!\!:\,
            $(\LET~\pz\ $\leftarrow$ (1~+~\py) \IN
            \UNL{7} \hspace*{1.15cm} $\psi_3\!\!:\, (\RETURN ~\pz)$))))))
        \end{uprogram}}
        \renewcommand{\arraystretch}{1}{
	  \begin{uprogram}
	  \UNL{1} $\pi_\mainpgm\!\!:\, \pi_8\!\!:\, $(\LET\  \pa\  $\leftarrow$ \ldots  \IN
	  \UNL{2}   \hspace*{.78cm}   $\pi_9\!\!:\,      $ (\LET\ \pb\  $\leftarrow$  \ldots \IN 
          \UNL{3}   \hspace*{.78cm}    $\pi_{10}\!\!:\,
          $(\LET\ \pw\  $\leftarrow$  (\length\ \pb) \IN
          \UNL{4}  \hspace*{.78cm}  $\psi_{4}\!\!:\,$ (\RETURN\ \pw)))))))
\end{uprogram}}
        } 
    \end{boxedminipage}
    & 
    \begin{boxedminipage}{.49\textwidth}
      {\sf
        \small
        \begin{tabular}{lc}
        $\cal{M_{\mbox{\tiny \length}}} = $ 
            \begin{tabular}{lc}            
              $[$ \\
              $\psi_1 \mapsto \lbrace \px.\sigma_{\length}, \xl.\cbar\sigma_{\length} \rbrace ,$ \\ 
              $\psi_2 \mapsto \lbrace \rbrace , $\\
              $\psi_3 \mapsto \lbrace \pz.\sigma_{\length}, \py.\cbar\sigma_{\length}, $\\
              $\pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}} \rbrace $  \\
              $]$
            \end{tabular} \\
            $\cal{P_{\mbox{\tiny \length}}} = $
              \begin{tabular}{lc}
                $[$ \\
                $\pi_{\mbox{1}} \mapsto \lbrace \psi_1, \psi_2, \psi_3 \rbrace, $ \
                $ \pi_{\mbox{2}} \mapsto \lbrace \psi_1, \psi_2, \psi_3 \rbrace, $ \\
                $ \psi_{\mbox{1}} \mapsto \lbrace \psi_1 \rbrace, $ \
                $ \psi_{\mbox{2}} \mapsto \lbrace \psi_2 \rbrace, $ \
                $ \pi_{\mbox{5}} \mapsto \lbrace \psi_3 \rbrace, $ \\
                $ \pi_{\mbox{6}} \mapsto \lbrace \psi_3 \rbrace, $ \
                $ \pi_{\mbox{7}} \mapsto \lbrace \psi_3 \rbrace $ \\
                $ \psi_{\mbox{3}} \mapsto \lbrace \psi_3 \rbrace $ \\
                $]$
              \end{tabular}\\
              $\cal{M_{\mbox{\tiny \main}}} = $ 
              \begin{tabular}{lc}            
                $[$
                  $\psi_4 \mapsto \lbrace \pw.\sigma_{\mathit{\!all}}, \pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace $ \
                  $]$
              \end{tabular} \\
              $\cal{P_{\mbox{\tiny \main}}} = $
                \begin{tabular}{lc}
                  $[$ \
                  $\pi_{\mbox{8}} \mapsto \lbrace \psi_4 \rbrace, $ \
                  $ \pi_{\mbox{9}} \mapsto \lbrace \psi_4 \rbrace, $ \
                  $ \pi_{\mbox{10}} \mapsto \lbrace \psi_4 \rbrace, $ \
                  $ \psi_{\mbox{4}} \mapsto \lbrace \psi_4 \rbrace, $ \
                  $ \ldots  $ \
                  $]$  
                \end{tabular}
        \end{tabular}
      }
    \end{boxedminipage}
    \\
    (a) Example program.&
    \renewcommand{\arraystretch}{.9}{\begin{tabular}[t]{ll}
        (b) {\em e-path} to liveness environment map \\
        (c) Prog pt to {\em e-paths } map
    \end{tabular}}
  \end{tabular}
\end{picture}
\kern -3ex
\vspace{0.5in}
\caption{Example Program and its liveness maps.}\label{fig:mot-example2}    
\end{figure*}

@@

Since   \pz\  is  $(1 + ~\py)$,  the   liveness  of   $\sigma$  for
\pz\ translates  into a demand  of $\sigma$ on $(1+~\py)$,  which in
turn  generates  a  liveness  of $\{\py.\epsilon\}$  at
$\pi_7$.  The liveness of \py\ should be killed at $\pi_6$ since it is
beyond the scope of the definition of \py.  However, this does not cause any imprecision
in the way of the GC marking  more cells as live, since \py\ is yet to
come into existence at $\pi_6$  and its liveness will not be consulted
by the garbage collector at this point.
In an eager language, \py\ would not have been live at $\psi_3$, the
point of  its last  use. In a  lazy language, however,  the definition
$\pz  \leftarrow (1+~\py)$,  does  not result  in  an evaluation  of
$(1+~\py)$; instead  a closure is created.   The evaluation actually
takes place while evaluating  $(\RETURN~\pz)$. Thus it is necessary to
record the liveness  of $\py$ at the program  point $\psi_3$, indeed at
every program  point from $\psi_3$ up  to the beginning  of the program
(since  we do  not kill  liveness).  We  thus regard  the body  of the
function as  a tree and consider paths  from the root of  this tree to
program  points  which  trigger  a evaluation\footnote{Program points
  which trigger evaluation are labelled with $\psi$}---\SIF\  conditions  and
\RETURN\  expressions.  We  call such  paths {\em
  e-paths}. \comment{Update the e-path labels}  For the
function \length, there are three such paths $\psi_1 = \pi_1, \pi_2, \pi_3,
\pi_4,  \pi_5$, $\psi_2  =  \pi_1, \pi_2,  \pi_3,  \pi_4, \pi_6,  \pi_7,
\pi_8$, and $\psi_3 = \pi_1, \pi_2, \pi_3, \pi_4, \pi_9$.

The analysis  consists of  two maps: The  map $\cal{M}$ maps  a e-path to a common liveness environment, and a second map $\cal{P}$ maps
a program  point to the set of  e-paths on which  the point lies.
If a  program  point happens to fall  on more than one  such path, then
the  liveness environment of  the program  point is  the variable-wise
union of  the liveness environments  of the individual paths.   In the
example program, ${\cal M}(\psi_1)= \ldots$, ${\cal M}(\psi_2)=\ldots$ and
${\cal M}(\psi_3)=  \ldots$.  Since  ${\cal P}(pi_4)$ is  $\lbrace \psi_1,
\psi_2, \psi_3 \rbrace$, the liveness  of $pi_4$ is \ldots.  In summary,
our propagation of liveness is flow-insensitive but path-sensitive.
In the sequel
we  shall  describe  how  to  obtain $\cal{M}$.   The  formulation  of
$\cal{P}$ is easy and we do not elaborate it any further.

@@

\section{Generating Liveness Equations}
The  program  in  Figure~\ref{fig:mot-example2}  serves as  a  running
example. Given  an arbitrary function $\mathit{f}$,  we now illustrate
how  to generate  the equations  describing the  demand transformation
\Lfonly$_\mathit{f}$.   Starting with a  symbolic demand  $\sigma$, we
determine   \Lfun{e_{\mathit{f}}}{\sigma}{\Lfonly}.    Of   particular
interest  is the liveness  at the  program point  at the  beginning of
$e_{\mathit{f}}$ (assumed  to be $\pi_1$).   We generate
equations     of     the     form    $\Lf{\mathit{f}}{i}{\sigma}     =
\Lanv{\pi_1}{x_i}$, where  $x_i$ is the  $i^{\mbox{\footnotesize th}}$
parameter of $\mathit{f}$. Doing this for \length, we get:
\begin{align*}
 \Lanv{\pi_1}{\xl} &= \cbar\sigma \cup \acdr\Lf{\length}{1}{\cbar\sigma}
\end{align*}
Which gives the following equation for \Lfone{\length}: 
  \begin{eqnarray*}
    \Lf{\length}{1}{\sigma}
    &=& \cbar\sigma \cup \acdr\Lf{\length}{1}{\cbar\sigma}
   \end{eqnarray*}

@@

As mentioned earlier\comment{Check if already mentioned, Not mentioned earlier}, each of  the liveness rules modifies a demand by
prefixing   it with   symbols   in   the   above   set,   and   therefore
$\Lf{f}{i}{\sigma}$ has the closed form:
\begin{eqnarray}
\label{eq:LF:DI}
  \Lf{f}{i}{\sigma} &=& \Df{f}{i}.\sigma
\end{eqnarray}
where \Df{f}{i} are sets of  strings over the alphabet $\lbrace \acar,
\acdr,\bcar, \bcdr, \cbar \rbrace$.   Substituting the guessed form in
the  equation  describing  \Lfonly$_{\mathit  f}$, and  factoring  out
$\sigma$,  we  get an  equation  for  \Df{f}{i} that is
independent of $\sigma$.  Any solution for \Df{f}{i} yields a solution
for \Lfonly$_{\mathit f}$.  Applied to \Lfonly$_{\mathit \length}$, we
get:
  \begin{align*}
    \Df{\length}{1} &= \lbrace\cbar\rbrace 
    \cup \acdr\Df{\length}{1}\cbar
  \end{align*}

@@

The expression $e_{\mainpgm}$  is a special case; we  assume it may be
called externally with demand $\sigma_\mainpgm = \{\acar,\acdr\}^\ast$
(denoted $\sigma_{\!all}$).  This is because any part of its value may
be used by the environment---for printing the result, for instance.
{\color {red} What about print?}
For the running example, $\length$ has calls from $\main$ at $\pi_{10}$
and a recursive call at $\pi_6$.
%\newcommand{\deltacall}[3]{\delta_{#1}({#2},{#3})}
So $\sigma_{\length} =
     \deltacall{\length}{\pi_{11}}{\main}  \cup \deltacall{length}{\pi_6}{\length}$.
Filling in  the values gives:

\begin{align*}
\sigma_{\length}    &=
 \sigma_{\!all}  ~\cup~ \acdr\Lf{length}{1}{\cbar\sigma_{\mathit{\length}}}
\end{align*}

For some of the program points in \length\  and \main,  this gives
{\color {red} Prasanna, please fill this correctly}. You could use the
set notation if the map notation spills over: 
\begin{align*}
\Lan{\pi_1} &= \lbrace  \px.\sigma_{\length}, \xl.\cbar\sigma_{\length}, \pz.\sigma_{\length},\
              \py.\cbar\sigma_{\length}, \pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}} \rbrace \\
\Lan{\pi_2} & =  \lbrace  \px.\sigma_{\length}, \xl.\cbar\sigma_{\length}, \pz.\sigma_{\length},\
                  \py.\cbar\sigma_{\length}, \pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}} \rbrace \\
\Lan{\pi_5} &= \lbrace \pu.\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}, \xl.\acdr\Lf{\length}{\mbox{1}}{\cbar\sigma_{\length}}\rbrace \\
\Lan{\pi_8} &= \lbrace \pw.\sigma_{\mathit{\!all}}, \pb.\Lf{\length}{\mbox{1}}{\sigma_{\mathit{\!all}}}, \ldots  \rbrace 
\end{align*}

In summary, the equations generated during liveness analysis are:
\begin{enumerate}
\item   For    each   function   $\mathit{f}$,    equations   defining
\Df{f}{i} for use by \Lfonly$_{\mathit f}$.
\item  For  each   function  $\mathit{f}$, an  equation  defining
the summary demand  $\sigma_{\mathit f}$ on $e_f$.
% \marginpar{Consistency: summary vs cumulative?}
\item For each function $\mathit{f}$ (including $\mainpgm$ for
  $e_\mainpgm$) \comment{Have we defined GC points for lazy languages? No, we define it only in the implementation section as of now.}
an equation defining
  liveness at each GC point of $e_{\mathit f}$.

\subsection{Solving liveness equations---the grammar interpretation}
The  equations above can now be re-interpreted as a
context-free grammar (CFG)  on the alphabet $\lbrace\acar, \acdr,
\bcar, \bcdr, \cbar\rbrace$.  Let  \var{$X$} denote the non-terminal for
a variable  $X$ occurring on  the LHS of the  equations generated
from the analysis.  We can  think of the resulting productions as
being associated with several  grammars, one for each non-terminal
\var{\Lanv{i}{x}} regarded as a start symbol.  As an example, the
grammar    for   \var{\Lanv{5}{\xl}}   comprises    the   following
productions:
 \begin{eqnarray*}
\var{\Lanv{5}{\xl}}  &\rightarrow&   \var{\acdr \Lf
  {\length}{1}{\cbar\sigma_{\length}}} \mid \cbar\sigma_{\length} \\
\end{eqnarray*}
\comment{Pick up a richer equation or more than one liveness equations}
Other equations can be converted  similarly.     The  language  generated  by  \var{\Lanv{i}{x}},
denoted $\mathscr{L}(\var{\Lanv{i}{x}})
$,  is the desired  solution of
\Lanv{i}{x}.   
However, recall from  our earlier  discussion that
the  decision problem that  we are  interested in  during garbage
collection is:
\begin{quote}
Let $x.\alpha$ be  a forward  access path  consisting  only  of   edges
\acar\  and \acdr\  (but  not \bcar\  or  \bcdr).   Let              $\mathscr{L}(\var{\Lanv{i}{x}})
\stackrel{*}{\hookrightarrow} \sigma$, where  $\sigma$  consists 
of forward paths only. Then does $\alpha \in \sigma$?
\end{quote}
We could convert  the  rules  defining  $\hookrightarrow$  into
productions and  add them to the grammar. However, this results
in  an {\em  unrestricted grammar}~\cite{hopcraft90toc},  and the
membership  problem   for  such  grammars   is
undecidable. \comment{Rewrite it to say that this is a CFG with 3
  unrestricted rules and the membership for even this grammar is undecidable}
\comment{State the undecidability of the decision problem above.}
